Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-12-14 prop_letter_ajhs.tex
done?

When combining non-disjoint events, there is a further subtlety whose
resolution will give us the final general condition for IVPMs. For
events $E_1$ and~$E_2$, not necessarily disjoint, we have:
\begin{equation}
\mu(E_1\cup E_2) + \mu(E_1\cap E_2) \subseteq \mu(E_1) + \mu(E_2)\,,
\label{eq:classicalconvex}
\end{equation}
which is a generalization of the classical inclusion-exclusion
principle that uses $\subseteq$ instead of $=$ for the same reason as
before. The new condition, known as \emph{convexity}
\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Marinacci1999,MarinacciMontrucchio2005,Grabisch2016},
reduces to the previously motivated Eq.~(\ref{eq:disjoint}) when the
events are disjoint, i.e., when $\mu(E_1\cap E_2) = 0$.

@@

\newpage

\begin{suggest}{\andy}{6}[inverted logic and phrasing.  I would start at "To formalize
this ..." and replace all the way to Eq.~(\ref{eq:hbar})]
We now show that our definition of QIVPMs is consistent
with classical IVPMs in the sense that a restriction of QIVPMs to
mutually commuting events recovers the definition of classical
IVPMs.  However, we must refine this context for the quantum case,
defining a QIVPM  $\bar{\mu}$ to be \emph{consistent} with a state $\rho$ and
a projector~$P$ if the interval $\bar{\mu}(P)$ contains the exact
probability calculated by the Born
rule~\citep{Born1983bibTeX,Mermin2007,Jaeger2007}, i.e., 
  \begin{equation}
  \muB_{\rho}\left(P\right) = \Tr\left(\rho
  P\right)\in\bar{\mu}\left(P\right)\ .
  \end{equation}
In contrast with conventional probability spaces, there is \emph{no
guarantee} that a unique state $\rho$ consistent with a QIVPM exists.

@@

\begin{definition}[The core of a probability measure] The \emph{core}
$\coreBorn\left(\bar{\mu},\events'\right)$ of a probability measure 
$\bar{\mu}$ relative to a subspace of events $\events'$ is the
collection of all states $\rho$ that are \emph{consistent} with 
$\bar{\mu}$ on every projector in $\events'$, that is
\begin{equation}
\label{eq:hbar}
\coreBorn\left(\bar{\mu},\events'\right) = \set{\rho}{\forall P\in
\events',~\muB_{\rho}\left(P\right)\in\bar{\mu}\left(P\right)} \ .
\end{equation}
\end{definition}
\end{suggest}
%% which coherently introduces the "core" and sets if off as a
%% DEFINITION, etc.

\newpage

We now show that our definition of QIVPMs is consistent with classical
IVPMs in the sense that a restriction of QIVPMs to mutually commuting
events recovers the definition of classical IVPMs. To formalize this
connection, we first note that in contrast to the case with
conventional quantum probability spaces, there is generally no unique
state~$\rho$ consistent with a QIVPM. We say that a QIVPM $\bar{\mu}$
is \emph{consistent} with a state $\rho$ and projector~$P$ if the
interval $\bar{\mu}(P)$ contains the exact probability calculated by
the Born rule, i.e, if:
\begin{equation}
\muB_{\rho}\left(P\right)\in\bar{\mu}\left(P\right)
\end{equation}
where $\muB_{\rho}\left(P\right)=\Tr\left(\rho P\right)$
is the probability of the event~$P$ as determined by the Born
rule~\citep{Born1983bibTeX,Mermin2007,Jaeger2007}.

We generalize this consistency relation to sets of states and general
subspaces of events as follows. A set of events $\events'$ is a
\emph{subspace} of $\events$ if it contains $\mathbb{0}$
and~$\mathbb{1}$ and is closed under complements, sums, and products,
i.e., for any projector~$P\in\events'$, we have
$\mathbb{1}-P\in\events'$ and for each pair of commuting
projectors~$P_{0}\in\events'$ and $P_{1}\in\events'$, we have
$P_{0}+P_{1}-P_{0}P_{1}$ and $P_{0}P_{1}\in\events'$. Given a
probability measure $\bar{\mu}$ and a subspace of events $\events'$,
we define the \emph{core} of $\bar{\mu}$ relative to $\events'$ as the
collection of all states $\rho$ that are consistent with $\bar{\mu}$
on every projector in $\events'$ as $\coreBorn(\bar{\mu},\events')$:
\begin{equation}
\label{eq:hbar}
\coreBorn\left(\bar{\mu},\events'\right)=\set{\rho}{\forall P\in \events',~\muB_{\rho}\left(P\right)\in\bar{\mu}\left(P\right)}.
\end{equation}

\newpage

\begin{suggest}{\andy}{7}[This story order is incomprehensible. I would state FIRST that 'here
is the outline of the proof', and LAST 'Thm 1..., whose full proof
will appear in~\cite{TaiThesis2018}.]
   The general idea of the proof that QIVPMs become classical
IVPMs when the space of events includes only mutually commuting
events depends on showing the existence of a non-empty core for such
commuting events.  The first step in establishing this assertion is to
to take a subspace $\eventsC$ of mutually commuting events, construct
a partial orthonormal basis by diagonalization, and complete this to a
full orthonormal basis  $\overline{\eventsC}$.  We then
build a bijection between the QIVPM on the set of projectors
associated with this basis and the set of classical events
corresponding to this basis.  Using this correspondence together with a
classical result by
Shapley~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Grabisch2016},
we can establish that for the special case of commuting events, a
QIVPM will always have a non-empty core~\footnote{Shapley's result
  proves that a classical IVPM always contains at least one state that
  is consistent with every event.}.  This leads to
\begin{thm}[Non-empty Core for Compatible Measurements] \label{thm:Shapley}
  For every QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$, if a subspace
  of events~$\eventsC\subseteq\events$ commutes, then
  $\coreBorn\left(\bar{\mu},\eventsC\right)\ne\emptyset$,
\end{thm}
whose full proof will appear in ~\cite{TaiThesis2018}.
\end{suggest}
The proof that QIVPMs become classical IVPMs when the space of events
includes only mutually commuting events is presented in detail in a
forthcoming thesis~\cite{TaiThesis2018}. The idea of the proof is as
follows. Given a subspace of mutually commuting events~$\eventsC$, we
can construct a partial orthonormal basis by diagonalization and
complete it to a full orthonormal basis $\overline{\eventsC}$. We then
build a bijection between the QIVPM on the set of projectors
associated with this basis and the set of classical events
corresponding to this basis. Using this correspondence together with a
classical result by
Shapley~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Grabisch2016},
we can establish that for the special case of commuting events, a
QIVPM will always have a non-empty core~\footnote{Shapley's result
  proves that a classical IVPM always contains at least one state that
  is consistent with every event.}.

@@

\begin{suggest}{\andy}{8}[inverted narrative order]
The importance of the existence of empty cores combined with
  the conditions guaranteeing non-empty cores is that one cannot
  guarantee non-empty cores for finite-precision attempts at proving
  Gleason's theorem by extending the Born measure
  $\muB_{\rho}\left(P\right)$ to QIVPMs $\bar{\mu}\left(P\right)$.  We
  will show in  Sec.~\ref{sec:Gleason} that a core defined by the set
  of states $\coreBorn\left(\bar{\mu},\events'\right)$ that is
  consistent with some arbitrary (not necessarily commuting)
  events~$\events'$  and a particular QIVPM  $\bar{\mu}$
   \emph{may be empty}.
\end{suggest}
The importance of this property is that it is violated in
finite-precision versions of Gleason's theorem
(cf. Thm.~\ref{thm:Non-extensible-of-Gleason's} in Sec.~\ref{sec:Gleason}). As we will show, the set of states
$\coreBorn\left(\bar{\mu},\events'\right)$ that is consistent with
some arbitrary (not necessarily commuting) events~$\events'$ and a
particular QIVPM $\bar{\mu}$ may be \emph{empty}.

>>>>

When combining non-disjoint events, there is a further subtlety whose
resolution will give us the final general condition for IVPMs. For
events $E_1$ and~$E_2$, not necessarily disjoint, we have:
\begin{equation}
\mu(E_1\cup E_2) + \mu(E_1\cap E_2) \subseteq \mu(E_1) + \mu(E_2)\,,
\label{eq:classicalconvex}
\end{equation}
which is a generalization of the classical inclusion-exclusion
principle that uses $\subseteq$ instead of $=$ for the same reason as
before. The new condition, known as
\emph{convexity}~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Marinacci1999,MarinacciMontrucchio2005,Grabisch2016},
reduces to the previously motivated Eq.~(\ref{eq:disjoint}) when the
events are disjoint, i.e., when $\mu(E_1\cap E_2) = 0$.

@@

Our definition of QIVPMs is consistent with classical IVPMs in the
sense that a restriction of QIVPMs to mutually commuting events,
$\eventsC$, recovers the definition of classical IVPMs. The proof of
this fact is included in the forthcoming thesis by the first author. A
consequence of this fact is that known properties of classical IVPMs
directly hold for QIVPMs when one restricts to mutually commuting
events, $\eventsC$. In particular, in the classical world, it is
\emph{impossible} for experiments to result in probabilities that are
inconsistent with \emph{some} state of the system under consideration,
i.e., all IVPMs must have a non-empty ``core''~\footnote{A result by
  Shapley~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Grabisch2016}
  proves that a classical IVPM always contains at least one state that
  is consistent with every event.}. Interestingly, as we show in
Sec.~\ref{sec:Gleason}, it is possible in the quantum world for the
probabilities associated with some events to be inconsistent with
\emph{any} quantum state, i.e., for the QIVPM to have an empty core;
in that case, one cannot guarantee non-empty cores for finite-precision
attempts at proving Gleason's theorem by extending the Born measure
$\muB_{\rho}\left(P\right)$ to QIVPMs
$\bar{\mu}\left(P\right)$. However, if the set of events is mutually
commuting, the situation reverts to the classical case in which
probabilities always determine at least one state.

We now give the necessary technical definitions to prove this
non-empty core property. We say a QIVPM $\bar{\mu}$ is
\emph{consistent} with a state $\rho$ and a projector~$P$ if the
interval $\bar{\mu}(P)$ contains the exact probability calculated by
the Born rule~\citep{Born1983bibTeX,Mermin2007,Jaeger2007}, i.e.,
  \begin{equation}
  \muB_{\rho}\left(P\right) = \Tr\left(\rho
  P\right)\in\bar{\mu}\left(P\right)\ .
  \end{equation}
In contrast with classical probability spaces, there is \emph{no
guarantee} that there exists a unique state $\rho$ consistent with a QIVPM.

@@

\begin{definition}[The core of a probability measure] The \emph{core}
$\coreBorn\left(\bar{\mu},\events'\right)$ of a probability measure 
$\bar{\mu}$ relative to a subspace of events $\events'$ is the
collection of all states $\rho$ that are \emph{consistent} with 
$\bar{\mu}$ on every projector in $\events'$, that is
\begin{equation}
\label{eq:hbar}
\coreBorn\left(\bar{\mu},\events'\right) = \set{\rho}{\forall P\in
\events',~\muB_{\rho}\left(P\right)\in\bar{\mu}\left(P\right)} \ .
\end{equation}
\end{definition}

We are now in position to state and prove that, for the special case
of commuting events, a QIVPM will always have a non-empty core.

@@

The proof, detailed in the forthcoming thesis, proceeds as
follows. From a subspace $\eventsC$ of mutually commuting events, one
can construct a partial orthonormal basis by diagonalization, and
complete this to a full orthonormal basis $\overline{\eventsC}$.  We
can then build a bijection between the QIVPM on the set of projectors
associated with this basis and the set of classical events
corresponding to this basis.  Using this correspondence together with
the classical result by
Shapley~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Grabisch2016},
we can establish that for the special case of commuting events, a
QIVPM will always have a non-empty core.
